Categorical Predictor Embeddings in Actuarial Modeling

### 2021-02-02 manuscript/manuscript.Rmd
clean up a bit

---
title: Categorical Predictor Embeddings in Actuarial Modeling
authors:
  - name: Kevin Kuo 
    affiliation: RStudio and Kasa AI
    email: kevin@kasa.ai
  - name: Ronald Richman
    affiliation: QED Actuaries and Consultants
    email: ronaldrichman@gmail.com
abstract: |
  Enter the text of your abstract here.
keywords:
  - blah
  - blee
  - bloo
  - these are optional and can be removed
bibliography: references.bib
output: rticles::arxiv_article 
extra_dependencies: ["amsmath"]

@@

Recently, several studies of insurance problems have applied an alternative approach, originally from the natural language processing literature [@Bengio2003] and now applied to diverse types of machine learning problem [@Guo2016], known as **categorical embeddings**. Instead of trying to capture the differences between categories using a single factor, the categorical embedding approach rather maps each category to a low-dimensional numeric vector, which is then used within the model as a new predictor variable.

@@

where the regression parameters (weights) for each layer $$k \in [1;K]$$ are represented by the matrices $a_k$ and the intercept terms are represented by $b_k$. Whereas the calculation inside each of the layers is nothing more than linear regression, $\sigma$ represents the non-linear part of each layer. Choices for $\sigma$ are often the tanh function or the rectified linear unit (ReLu) $max(0,x)$. The parameters of the network are estimated ('trained') as follows. First a loss function $L(.,.)$ is specified for the network that measures the difference between the observed data $y$ and the predictions of the network $\hat{y}$, for example, the Mean Squared Error ($(y - \hat{y})^2$). Then, the parameters of the network are changed such that the loss decreases (formally, this is done using the technique of backpropagation).Finally, training is stopped once the predictive performance of the network on unseen data is suitably good.

@@

### Embeddings
Common issues with the traditional encoding schemes for categorical data occur when the number of levels for each variable is very large. Often, in these cases, models do not converge quickly, and the very large matrices that result from applying these schemes often cause computational difficulties. Besides for these practical issues, a deeper issue is that one-hot or dummy encoded data assumes that each category is entirely independent of the rest of the categories, in other words, there are no similarities between categories that could enable more robust estimation of models. In technical terms, this is because the columns of the matrices created by one-hot encoding are all orthogonal to each other. (These arguments appear in a similar form in [@Guo2016].) Solutions to these problems are provided by embedding layers.

@@

where $q_\mathcal{P^j}$ is the dimension of the embedding for the $j$th categorical variable and $z_\mathcal{P^j}(.)$ is an function that maps from the particular element of the labels $p$ to the embedding space. Equation \ref{eq:embedding} states that an embedding maps a level of a categrical variable to a numerical vector. This function is left implicit, meaning to say, we allow the embeddings to be derived during the process of fitting the model and do not attempt to specify exactly how the embeddings can be derived from the input data. In Table \ref{embed} we show an example of two dimensional embeddings for the state variables, where these have been generated randomly.

```{r, echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
states= fread("../manuscript/states.csv")
set.seed(42)

data.frame(sample_dat, dimension1 = rnorm(5),dimension2 = rnorm(5)) %>% knitr::kable(label = "", caption = "Example (random) embeddings of the state variable")

@@

In this section, we describe, fit, and evaluate both a minimalist neural network architecture utilizing embeddings and a traditional GLM. The working example for our experiments is as follows: Given a set of claims characteristics, we predict the losses paid on the property coverage of the policy. In the rest of this section, we describe the dataset we use, describe formally the models being considered, and discuss results.

@@

The data we use comes from the National Flood Insurance Program (NFIP) and is made available by the OpenFEMA initiative of the Federal Emergency Management Agency (FEMA) (TODO: cite). Two datasets are made available by OpenFEMA: A policies dataset with exposure information, and a claims dataset with claims transactions, including paid amounts. Because there is no way to associate records of the two datasets, we are limited to fitting severity models on the claims dataset. While the complete dataset contains over two million transactions, for the purposes of our experiments we limit ourselves to data from 2000 to August 2019, which amounts to approximately 1.4 million claims. The dataset can be downloaded from Cellar (TODO: link). The dataset includes a rich variety of variables, from occupancy type to coarse coordinates. For our models, we work with a few selected variables that represent continuous and discrete variables of low and high cardinalities, which we list in Table \ref{tab:variables}.

| Variable                                 | Type                                     |
|------------------------------------------|------------------------------------------|
| Building insurance coverage              | Numeric                                  |
| Basement enclosure type                  | Categorical                              |
| Number of floors in the insured building | Categorical (binned in original dataset) |
| Flood zone                               | Categorical (high cardinality)           |
| Primary residence                        | Categorical (binary indicator)           |

: \label{tab:variables} Variables used in modeling.

@@

We fit a GLM with the following specifications:

-   Target variable: Amount paid on building claim

-   Predictors: *Log* of **building insurance coverage**, **basement enclosure type**, **number of floors in the insured building**, *prefix* of **flood zone**, and ***primary residence***.

-   Link function: log

-   Distribution: gamma

We take the log of the continuous predictor **building insurance coverage** following (cite cas glm monograph), which allows the scale of the predictor to match that of the target variable. Because the **flood zone** variable in the original data contains 60 levels, we take the prefix of the zone code, which corresponds to the level of risk as determined by FEMA (source?). For example, *A01*, *A02*, and so on are recoded as simply *A*.

@@

(Note: this section is WIP doesn't match up w/ code, which is also WIP) For the neural net, we utilize the same responsible variable, amount paid on building claim. We normalize the numeric variable ***building insurance coverage***, one-hot encode the **primary residence** variable, and apply embedding layers of dimension one to the rest of the categorical variables. We then connect the embeddings and the normalized variable to a single output unit with softplus activation, where the softplus function is defined to be

\\begin{equation} x \\mapsto \\log(1 + e\^x). \\end{equation}

Figure (todo: add figure) exhibits the architecture of our simple neural network.

>>>>

---
title: Categorical Predictor Embeddings in Actuarial Modeling
authors:
  - name: Kevin Kuo 
    affiliation: RStudio and Kasa AI
    email: kevin@kasa.ai
  - name: Ronald Richman
    affiliation: QED Actuaries and Consultants
    email: ronaldrichman@gmail.com
abstract: |
  Enter the text of your abstract here.
keywords:
  - blah
  - blee
  - bloo
  - these are optional and can be removed
bibliography: references.bib
output:
  bookdown::pdf_book:
    base_format: rticles::arxiv_article
extra_dependencies: ["amsmath"]

@@

Recently, several studies of insurance problems have applied an alternative approach, originally from the natural 
known as **categorical embeddings**. Instead of trying to capture the differences between categories using a single 
factor, the categorical embedding approach rather maps each category to a low-dimensional numeric vector, which is then
used within the model as a new predictor variable.

@@

where the regression parameters (weights) for each layer $$k \in [1;K]$$ are represented by the matrices $a_k$ and the 
intercept terms are represented by $b_k$. Whereas the calculation inside each of the layers is nothing more than linear 
regression, $\sigma$ represents the non-linear part of each layer. Choices for $\sigma$ are often the tanh function or 
the rectified linear unit (ReLu) $max(0,x)$. The parameters of the network are estimated ('trained') as follows. First a 
loss function $L(.,.)$ is specified for the network that measures the difference between the observed data $y$ and the 
predictions of the network $\hat{y}$, for example, the Mean Squared Error ($(y - \hat{y})^2$). Then, the parameters of 
the network are changed such that the loss decreases (formally, this is done using the technique of backpropagation).
Finally, training is stopped once the predictive performance of the network on unseen data is suitably good.

@@

### Embeddings

Common issues with the traditional encoding schemes for categorical data occur when the number of levels for each 
variable is very large. Often, in these cases, models do not converge quickly, and the very large matrices that result 
from applying these schemes often cause computational difficulties. Besides for these practical issues, a deeper issue 
is that one-hot or dummy encoded data assumes that each category is entirely independent of the rest of the categories, 
in other words, there are no similarities between categories that could enable more robust estimation of models. In 
technical terms, this is because the columns of the matrices created by one-hot encoding are all orthogonal to each 
other. (These arguments appear in a similar form in [@Guo2016].) Solutions to these problems are provided by embedding 
layers.

@@

where $q_\mathcal{P^j}$ is the dimension of the embedding for the $j$th categorical variable and $z_\mathcal{P^j}(.)$ is 
a function that maps from the particular element of the labels $p$ to the embedding space. Equation \ref{eq:embedding} 
states that an embedding maps a level of a categrical variable to a numerical vector. This function is left implicit, 
meaning to say, we allow the embeddings to be derived during the process of fitting the model and do not attempt to 
specify exactly how the embeddings can be derived from the input data. In Table \ref{embed} we show an example of two 
dimensional embeddings for the state variables, where these have been generated randomly.

```{r, echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
states <- fread("../manuscript/states.csv")
set.seed(42)

sample_dat <- sample_n(states, 5) %>%
  arrange(state)

sample_dat <- sample_n(states, 5) %>%
  arrange(state)

data.frame(sample_dat, dimension1 = rnorm(5), dimension2 = rnorm(5)) %>%
  knitr::kable(label = "", caption = "Example (random) embeddings of the state variable")
```

@@

In this section, we describe, fit, and evaluate both a GLM and a minimalist neural network architecture utilizing
embeddings. While this paper focuses on embeddings in neural networks, we begin with a GLM to provide a common frame of 
reference since most actuaries are familiar with the technique. The working example for our experiments is as follows: 
Given a set of claims characteristics, we predict the losses paid on the property coverage of the policy. In the rest of 
this section, we describe the dataset we use, describe formally the models being considered, and discuss results.

@@

The data we use comes from the National Flood Insurance Program (NFIP) and is made available by the OpenFEMA initiative 
of the Federal Emergency Management Agency (FEMA)[^9]. Two datasets are made available by OpenFEMA: A policies dataset 
with exposure information, and a claims dataset with claims transactions, including paid amounts. Because there is no 
way to associate records of the two datasets, we are limited to fitting severity models on the claims dataset. While the 
complete dataset contains over two million transactions, for the purposes of our experiments we limit ourselves to data 
from 2000 to August 2019, which amounts to approximately 1.4 million claims. The dataset can be downloaded from 
Cellar[^10] The dataset includes a rich variety of variables, from occupancy type to coarse coordinates. For our models, 
we work with a few selected variables that represent continuous and discrete variables of low and high cardinalities, 
which we list in Table \@ref(tab:variables)

[^9]: https://www.fema.gov/about/reports-and-data/openfema
[^10]: https://cellar.kasa.ai

```{r variables, results='asis', echo = FALSE}
library(magrittr)
variables <- tibble::tribble(
  ~Variable, ~Type,
  "Building insurance coverage", "Numeric",
  "Basement enclosure type", "Categorical",
  "Number of floors in the insured building", "Categorical (binned in original dataset)",
  "Flood zone", "Categorical (high cardinality)",
  "Primary residence", "Categorical (binary indicator)"
)
knitr::kable(
  variables,
  caption = "Predicator variables.",
  booktabs = TRUE,
  table.env = "table",
  escape = FALSE,
  linesep = ""
) %>%
  kableExtra::kable_styling(latex_options = "hold_position")
```

@@

While GLMs are commonplace and well-studied in the actuarial literature, there are still plenty of decisions to be made
in the modeling process; see, for example, @Goldburd for an in-depth discussion. For our purpose of establishing a
baseline model, we proceed with what we perceive as reasonable decisions around feature engineering and model structure, outlined below:

-   Target variable: Amount paid on building claim
-   Predictors: *Log* of **building insurance coverage**, **basement enclosure type**, **number of floors in the insured building**, *prefix* of **flood zone**, and ***primary residence***.
-   Link function: log
-   Distribution: gamma

We take the log of the continuous predictor **building insurance coverage**, which allows the scale of the predictor to 
match that of the target variable. Because the **flood zone** variable in the original data contains 60 levels, we take 
the prefix of the zone code, which corresponds to the level of risk as determined by FEMA [^8]. For example, *A01*, 
*A02*, and so on are recoded as simply *A*.

[^8]: https://www.floodsmart.gov/flood-map-zone/find-yours

@@

For the neural net, we utilize the same response variable, amount paid on building claim. We perform the following 
transformations on the predictor variables:

- Normalize the numeric variable ***building insurance coverage*** (center by mean and scale by standard deviation), 
- One-hot encode the **primary residence** variable, and 
- Apply embedding layers of dimension one to the rest of the categorical variables.

In terms of architecture, we connect the embeddings and the normalized numeric variable to a single output unit with 
softplus activation, where the softplus function is defined to be

$$
x \mapsto \log(1 + e^x). 
$$

Figure \@ref(fig:nn-1) exhibits the architecture of our simple neural network.

```{r nn-1, echo = FALSE, fig.cap = "blahblah", out.width = "80%", fig.align = "center"}
knitr::include_graphics("images/nn.png")
```
